---
title: "Compare Models And Select The Best Using The Caret R Package"
author: "Mark Goldberg"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
#weight: 1
draft: false
math: true
tags: ["R", "Statistics", "caret", "Machine Learning"]
categories: ["Statistics"]
#bibliography: [bib.bib]
output:
  blogdown::html_page:
    toc: true
summary: "Compare Models And Select The Best Using The Caret R Package."
---

## Compare Models And Select The Best Using The Caret R Package

*Data*: mlbench::PimaIndiansDiabetes. Find the best model to predict diabetes from all given parameters.  

*Models*:  
* **Learning Vector Quantization** (LVQ)  
* **Gradient Boosted Machine** (GBM)  
* **Support Vector Machine** (SVM)  

Each model is automatically tuned and is evaluated using 3 repeats of 10-fold cross validation.  
The random number seed is set before each algorithm is trained to ensure that each algorithm gets the same data partitions and repeats.  
The best models have 30 results (3 repeats of 10-fold cross validation).  
The objective of comparing results is to compare the accuracy distributions (30 values) between the models.  

This is done in three ways. The distributions are summarized in terms of the percentiles. The distributions are summarized as box plots and finally the distributions are summarized as dot plots.  

```{r}
library(mlbench)
library(caret)

# load the dataset
data(PimaIndiansDiabetes)

# training scheme
control <- trainControl(method="repeatedcv", number=10, repeats=3)

# train the LVQ model
set.seed(7)
modelLvq <- train(diabetes~., data=PimaIndiansDiabetes, method="lvq", trControl=control)

# train the GBM model
set.seed(7)
modelGbm <- train(diabetes~., data=PimaIndiansDiabetes, method="gbm", trControl=control, verbose=FALSE)

# train the SVM model
set.seed(7)
modelSvm <- train(diabetes~., data=PimaIndiansDiabetes, method="svmRadial", trControl=control)

# collect resamples
results <- resamples(list(LVQ=modelLvq, GBM=modelGbm, SVM=modelSvm))

# summarize the distributions
summary(results)
```
```{r}
# boxplots of results
bwplot(results)
```
```{r}
# dot plots of results
dotplot(results)
```

## Bibliography
[Compare Models And Select The Best Using The Caret R Package](https://machinelearningmastery.com/compare-models-and-select-the-best-using-the-caret-r-package/)