---
title: "How to split data into train and test subsets?"
subtitle: ""
author: "Mark Goldberg"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
lastmod: '`r format(Sys.time(), "%Y-%m-%d")`'
draft: false
math: true
tags: ["R", "Statistics"]
categories: ["R", "Statistics"]
#bibliography: [bib.bib]
output:
  toc: true
summary: "Here you will learn approaches to split your data into subsets - train and test for your modeling."
---
  
Here you can find several simple approaches to split data into train and test subset to fit and to test parameters of your model. 
We want to split our data:  
train percent - 0.7  
test percent - 0.3  

Let us create a data frame filled with generated data.  

```{r}
df = data.frame(x=seq(.1,1,.1), y=seq(1,10), z=letters[seq(1,10)])
df
```
1. First approach is to create a vector filled with selected IDs of rows and then apply this vector to subset data.  

```{r}
set.seed(1) # set state of random number generator
tv1 = sample(nrow(df), nrow(df)*0.7)
tv1

# Select train and test data for column y using our vector
train = df$y[tv1]
test = df$y[-tv1] # the same as test = df$y[!tv1]
train
test
```

2. Another approch is to create a vecotr filled with logical true/false for each row of dataset and apply this vector to subset data.  
```{r}
set.seed(1) # set state of random number generator
tv2 = sample(c(TRUE, FALSE), nrow(df), replace = T, prob = c(0.7,0.3))
tv2

# Select train and test data for column y using our vector
train = df$y[tv2]
test = df$y[!tv2]
train
test
```

3. Using [`caTools`](https://cran.r-project.org/web/packages/caTools/index.html).
```{r}
library(caTools)
set.seed(1) 
sample = sample.split(df, SplitRatio = .7)
train = subset(df$y, sample == TRUE)
test  = subset(df$y, sample == FALSE)
train
test
```

4. Using [`dplyr`](https://cran.r-project.org/web/packages/dplyr/)
```{r, message=F}
library(dplyr)
df$id <- 1:nrow(df)
train <- df %>% dplyr::sample_frac(.7)
test  <- dplyr::anti_join(df, train, by = 'id')
train$y
test$y
```