---
title: "Exploratory analysis in R"
author: "Mark Goldberg"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
draft: false
math: true
tags: ["R", "Statistics", "Exploratory analysis"]
categories: ["Statistics"]
#bibliography: [bib.bib]
output:
  #blogdown::html_page:
  toc: true
summary: "Exploratory analysis in R."
---

## Example 1: Iris
**Data description**: Iris is a famous (Fisher's or Anderson's) data set gives the measurements in centimeters of the variables **sepal** length and width and **petal** length and width, respectively, for 50 flowers from each of 3 species of iris. The species are Iris **setosa**, **versicolor**, and **virginica**.  

## Structure of the data
```{r}
# select numerical columns 1-4
df = iris[1:4]

head(df, n=5) # first rows
tail(df, n=5) # last rows
dim(df)       # data dimention
nrow(df)      # number of rows
ncol(df)      # number of columns
str(df)  
```

## Correlation between variables

Let's find if data in columns are correlated using `corrplot`.  
[Here](http://www.sthda.com/english/wiki/visualize-correlation-matrix-using-correlogram) you can find how to tune correlogram.  
```{r}
library(corrplot)

# build correlation matrix 
cor.mx <- cor(df)
# correlation plot ordered correlation coefficients
corrplot::corrplot.mixed(cor.mx, order="hclust")
```
```{r, message=F}
print(cor.mx)

library(caret)
# prepare training scheme
control <- caret::trainControl(method="repeatedcv", number=10, repeats=3)
#model <- train(diabetes~., data=PimaIndiansDiabetes, method="lvq", preProcess="scale", trControl=control)
# estimate variable importance
#importance <- varImp(model, scale=FALSE)
# summarize importance
#print(importance)
```
How to show correlation table? Alternatively we can use `corrplot` with `method = "number"`.   

Missing data
How to remove 'NA', 'Inf', and '0'
```{r}
colSums(sapply(df, is.na))
```

```{r}
# Scatterplot Matrices
pairs(iris[, -5], bg = c("yellow", "green", "black")[iris$Species], pch = 21)
```
How to choose variables with the best correlation? Can we sort them by importance?

Can we exclude data based on low correlation?
This can be also done by dimentionality reduction methods such as [PCA](/posts/statistics/pca/pca/).  

Can we remove outliers

Do we need to remove duplicated data (rows, columns)

# Filtering
You can see how to filter and subset your data in folloing posts:
1.  
2.  
3.  