---
title: "Exploratory analysis in R"
author: "Mark Goldberg"
date: '`r format(Sys.time(), "%Y-%m-%d")`'
draft: false
math: true
tags: ["R", "Statistics", "Exploratory analysis"]
categories: ["Statistics"]
#bibliography: [bib.bib]
output:
  #blogdown::html_page:
  toc: true
summary: "Exploratory analysis in R."
---
## View data
Here you can find several R functions which can help to see the structure of your data.  


## Correlation between variables

Let's find if data in columns are correlated using `corrplot`.  
[Here](http://www.sthda.com/english/wiki/visualize-correlation-matrix-using-correlogram) you can find how to tune correlogram.  
```{r}
library(corrplot)

# select numerical columns 1-4
df = iris[1:4]

# build correlation matrix 
cor.mx <- cor(df)

# correlation plot ordered correlation coefficients
corrplot::corrplot.mixed(cor.mx, order="hclust")
               
```
```{r, message=F}
print(cor.mx)

library(caret)
# prepare training scheme
control <- caret::trainControl(method="repeatedcv", number=10, repeats=3)
#model <- train(diabetes~., data=PimaIndiansDiabetes, method="lvq", preProcess="scale", trControl=control)
# estimate variable importance
#importance <- varImp(model, scale=FALSE)
# summarize importance
#print(importance)
```
How to show correlation table? Alternatively we can use `corrplot` with `method = "number"`.   


How to choose variables with the best correlation? Can we sort them by importance?

Can we exclude data based on low correlation?
This can be also done by dimentionality reduction methods such as [PCA](/posts/statistics/pca/pca/).  

Can we remove outliers

How to remove 'NA', 'Inf', and '0' 

Do we need to remove duplicated data (rows, columns)

# Filtering
You can see how to filter and subset your data in folloing posts:
1.  
2.  
3.  